\section{Operation}

This section describes the standard operation of automated image acquisition.
In normal use, the scheduler runs continuously as a background service on the
Raspberry Pi. Users control acquisition by creating or updating the schedule
file. The scheduler can be started before a schedule exists. If no schedule
file is present, it remains active and waits until a valid schedule is created.
During operation, it periodically checks whether the schedule file has changed.
If a valid new schedule is detected, the scheduler reloads and starts using the
new schedule.

\subsection{Schedule File}

The acquisition schedule is stored in:

\begin{verbatim}
$PHENOPI_ROOT/runtime/schedule.json
\end{verbatim}

This file defines the experiment start date, the number of experiment days, the
daily capture times, and optional technical replicate settings. A typical
schedule file is shown below.

\begin{verbatim}
{
  "start_date": "2026-05-01",
  "num_days": 14,
  "replicates": 3,
  "replicate_interval_seconds": 30
  "times": [
    "08:00",
    "08:30",
    "09:00",
    "09:30",
    "10:00"
  ],
}
\end{verbatim}

The \texttt{start\_date} field defines the first day of imaging. The
\texttt{num\_days} field defines the number of experiment days. The
\texttt{times} field defines the daily imaging time points in 24-hour
\texttt{HH:MM} format. The \texttt{replicates} field defines how many images
are acquired at each time point. The \texttt{replicate\_interval\_seconds}
field defines the delay between these replicate captures. For example, with
three replicates and a 30-second interval, captures scheduled at \texttt{08:00}
are taken at approximately \texttt{08:00:00}, \texttt{08:00:30}, and
\texttt{08:01:00}.

\subsection{Generating a Schedule}

Schedules can be written manually, but it is usually safer and more convenient
to generate them using the schedule generator. The generator creates a valid
\texttt{schedule.json} file and checks for overlapping capture times. The
schedule generator provides three ways to define daily capture times. The
\texttt{every} mode generates times between a start and end time, using a fixed
interval. This is useful for regular imaging throughout a photoperiod. The
\texttt{duration} mode generates times from a start time for a fixed duration.
This is useful when the length of the imaging window is easier to define than
the exact end time. The \texttt{centered} mode generates times around a central
time point. This is useful for imaging before and after a specific event, such
as treatment application.

For example, the following command creates a 14-day schedule with imaging every
30 minutes between 08:00 and 19:30. At each time point, three replicate images
are captured 30 seconds apart.

\begin{verbatim}
python -m scripts.scheduling.schedule_generator_cli every \
  --start 08:00 \
  --end 19:30 \
  --step-minutes 30 \
  --start-date 2026-05-01 \
  --num-days 14 \
  --replicates 3 \
  --replicate-interval-seconds 30 \
  --output runtime/schedule.json \
  --overwrite
\end{verbatim}

The generated file should be inspected before starting a long experiment. Check
that the start date, experiment duration, capture times, and replicate settings
are correct.

\subsection{Schedule Changes}

The scheduler periodically checks whether \texttt{runtime/schedule.json} has
changed. If the file changes and the new schedule is valid, the scheduler
reloads and uses the new schedule. If the file is missing or invalid, the
scheduler keeps the current schedule and reports the problem in the log. Each
schedule is associated with a hash-based scheduler database in the runtime
directory. This allows the scheduler to resume pending jobs after a restart
when the schedule has not changed. If the schedule file contents change, a new
database is used for the new schedule. Stale scheduler databases from previous
schedules are removed automatically.

\subsection{Captured Image Output}

All acquired images are stored in the configured capture directory. In the default installation, this is:

\begin{verbatim}
$PHENOPI_ROOT/captures
\end{verbatim}

Images are saved using the filename format:

\begin{verbatim}
capture_YYYYMMDD_HHMMSS.jpg
\end{verbatim}

For example:

\begin{verbatim}
capture_20260504_153945.jpg
capture_20260504_154000.jpg
capture_20260504_154015.jpg
\end{verbatim}

The timestamp in the filename records the acquisition time and is used by
downstream analysis. Users should therefore avoid renaming captured images.

\subsection{Monitoring an Experiment}

It is recommended to inspect the first few captures of each experiment. Check
that images are correctly exposed, focused, and contain the complete plant
tray. During the experiment, routine monitoring usually consists of checking
the service log, checking that new images are appearing in the capture
directory, and checking that enough storage space remains available. Recent
captured images can be listed with:

\begin{verbatim}
ls -lt "$PHENOPI_CAPTURE_DIR" | head
\end{verbatim}

Available storage space can be checked with:

\begin{verbatim}
df -h
\end{verbatim}

\subsection{Restarting or Stopping Acquisition}

The scheduler can be restarted with:

\begin{verbatim}
sudo systemctl restart phenopi-scheduler.service
\end{verbatim}

A restart is useful after updating the code or changing service-level settings.
If the schedule file has not changed, the scheduler resumes using the matching
schedule database. The scheduler can be stopped with:

\begin{verbatim}
sudo systemctl stop phenopi-scheduler.service
\end{verbatim}

Stopping the scheduler prevents new captures from being started. It does not
remove captured images.

\subsection{After Experiment Completion}

After the final scheduled capture has completed, the capture directory contains
the raw image dataset for the experiment. The capture directory should be
preserved unchanged. Downstream Phenopi analysis tools operate on these
timestamped images and write processed outputs to separate result directories.
Keeping the original images unchanged ensures that analyses can be repeated
later with updated settings or newer versions of the analysis pipeline.
