%-------------------------------------------------------------------------------
%    DOCUMENT CONFIGURATION
%-------------------------------------------------------------------------------

\documentclass[
    a4paper, % Paper size
    11pt, % Default font size
    twoside, % Change headers and footers between even and odd pages
]{JournalArticle} % Use the custom JournalArticle class

\raggedbottom%

\begin{document}

%-------------------------------------------------------------------------------
%    TITLE SECTION
%-------------------------------------------------------------------------------

\include{frontpage}

\newpage

%-------------------------------------------------------------------------------
%    TABLE OF CONTENTS
%-------------------------------------------------------------------------------

\parskip=0em
\renewcommand{\contentsname}{\centering \Large Contents}
\tableofcontents
\parskip=1em
\newpage

%-------------------------------------------------------------------------------
%    ARTICLE CONTENTS
%-------------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\subsection{Purpose}

Phenopi is an image-based plant phenotyping system for controlled-environment
experiments. It is designed for repeated RGB imaging of plants and for
extracting quantitative traits such as canopy area and growth over time. The
system combines automated image acquisition, image analysis, quality control,
and growth analysis in a modular workflow. It was developed as a low-cost
alternative to commercial phenotyping platforms and can run on inexpensive
hardware such as a Raspberry Pi camera setup. Users do not need extensive
programming experience, but Phenopi is currently command-line based and assumes
basic familiarity with running terminal commands and working with text-based
configuration files.

\subsection{What Phenopi Does}

Phenopi supports a complete basic workflow for image-based plant phenotyping.
During an experiment, a Raspberry Pi captures images at scheduled time points
and stores them using timestamped filenames. After acquisition, the analysis
pipeline processes the images, segments plant canopies, extracts traits, and
writes the results to output tables. The system also produces diagnostic
outputs that help users check image quality and segmentation accuracy before
interpreting the biological results. Growth analysis tools can then be used to
summarize canopy development over time.

\subsection{What This Manual Covers}

This manual explains how to install Phenopi for automated image acquisition,
prepare an experiment, run the scheduler, process captured images, inspect
quality-control outputs, and perform basic growth analysis. The manual focuses
on practical use. It describes the files users need to edit, the commands they
need to run, the outputs they should expect, and the checks they should perform
when something goes wrong. Detailed implementation notes are kept to a minimum.
Where internal behaviour is described, it is included only when it helps users
understand how to operate or troubleshoot the system.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{System Layout}

\subsection{Overview}

Phenopi is organized as a workflow rather than as a single application. The
main steps are image acquisition, image analysis, quality control, and growth
analysis. Each step reads input files from the experiment directory and writes
its own outputs, so intermediate results can be inspected before continuing. A
typical workflow starts with an experiment schedule. The Raspberry Pi uses this
schedule to capture images and store them in the image archive. The analysis
scripts then process these images, extract plant traits, and write the results
to tables. Diagnostic plots and overlays are used to check whether the images
and segmentations are suitable for downstream analysis.

\subsection{Hardware Components}

The acquisition setup consists of a Raspberry Pi, a Raspberry Pi Camera Module
v3, a fixed imaging frame, an experimental tray, and a stable light source. The
camera is mounted above the tray to acquire top-down RGB images. The hardware
setup should remain fixed during an experiment. Moving the camera, tray, or
light source can change the image geometry or illumination, which may affect
segmentation and trait extraction.

\subsection{Software Components}

Phenopi uses separate scripts for scheduling, image capture, image analysis,
plotting, and growth analysis. The scheduler runs on the Raspberry Pi and
starts capture jobs at the configured times. The capture script acquires
individual images and saves them using timestamped filenames. The analysis
scripts process the captured images after acquisition. They assign regions of
interest, segment plant canopies, extract traits, and save the results.
Plotting and growth analysis scripts use these trait tables to generate figures
and summary files.

\subsection{Directory Structure}

A typical Phenopi project or experiment directory contains folders for captured
images, runtime files, scripts, and results. The exact contents may differ
between installations, but the main folders should remain easy to identify.

\begin{forest}
  for tree={
    font=\ttfamily,
    grow'=0,
    child anchor=west,
    parent anchor=south,
    anchor=west,
    calign=first,
    edge path={
    \noexpand\path[draw, \forestoption{edge}]
    (!u.south west) + (7.5pt,0) |- (.child anchor)\forestoption{edge label};
    },
    before typesetting nodes={
    if n=1
    {insert before={[,phantom]}}
    {}
    },
    fit=band,
    before computing xy={l=15pt},
  }
  [phenopi/
    [captures/
      [capture\_20260508\_080000.jpg]
      [capture\_20260508\_080030.jpg]
      [\ldots]
    ]
    [deploy/
      [phenopi-scheduler.service]
    ]
    [results/]
    [runtime/
      [completed\_jobs.json]
      [schedule.json]
    ]
    [scripts/
      [analysis/]
      [capture/]
      [plotting/]
      [scheduling/]
    ]
  ]
\end{forest}

The most important configuration file for image acquisition is
\texttt{runtime/schedule.json}. This file defines when images should be
captured. The scheduler also writes its state to
\texttt{runtime/completed\_jobs.json}, which records which captures have
already been completed. Raw images are saved in \texttt{captures/} using
timestamped filenames. Analysis results are saved in \texttt{results/}, usually
as CSV files, diagnostic images, and plots. These files are the main outputs
used for quality control and biological interpretation. The \texttt{scripts/}
directory contains the command-line tools used for acquisition and analysis.
The \texttt{deploy/} directory contains the systemd service file used for
automated acquisition on the Raspberry Pi. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Installation on the Raspberry Pi}

\subsection{System Requirements}

Automated image acquisition requires a Raspberry Pi 4 or newer, a Raspberry Pi
Camera Module v3, Raspberry Pi OS Bookworm or newer, and Picamera2. The
scheduler is installed as a systemd service, so the Raspberry Pi should be set
up with a working network connection, correct system time, and enough storage
space for the planned experiment.

The installation below assumes that Phenopi will run under a dedicated user
account named \texttt{phenopi}. This keeps the acquisition service separate
from personal user accounts and makes file permissions easier to manage.

\subsection{Expected Installation Layout}

The provided service file assumes the following installation layout:

\begin{verbatim}
  /home/phenopi/phenotyping
  /home/phenopi/venvs/phenopi
\end{verbatim}

The Phenopi repository is stored in \texttt{/home/phenopi/phenotyping}. The
Python virtual environment is stored in \texttt{/home/phenopi/venvs/phenopi}.
If different paths are used, ensure to update the paths in
\texttt{deploy/phenopi-scheduler.service} also.

\subsection{Creating the Phenopi User}

Create a dedicated user for running the scheduler.

\begin{verbatim}
sudo useradd -m -s /bin/bash phenopi
\end{verbatim}

Create the directory that will contain the Python virtual environment.

\begin{verbatim}
sudo -u phenopi mkdir -p /home/phenopi/venvs
\end{verbatim}

\subsection{Installing Phenopi}

Clone the Phenopi repository into the expected location.

\begin{verbatim}
sudo -u phenopi git clone https://github.com/kfreinders/phenopi.git 
/home/phenopi/phenotyping
\end{verbatim}

Install Picamera2 using the Raspberry Pi package manager.

\begin{verbatim}
sudo apt update
sudo apt install -y python3-picamera2
\end{verbatim}

Create the Python virtual environment. The \texttt{--system-site-packages}
option makes packages installed through \texttt{apt}, such as Picamera2,
available inside the virtual environment.

\begin{verbatim}
sudo -u phenopi python3 -m venv --system-site-packages 
/home/phenopi/venvs/phenopi
\end{verbatim}

Install the Python dependencies listed by the project.

\begin{verbatim}
sudo -u phenopi /home/phenopi/venvs/phenopi/bin/pip install -r 
/home/phenopi/phenotyping/requirements.txt
\end{verbatim}

\subsection{Checking the Scheduler Command}

Before installing the service, check that the scheduler can be started from the
virtual environment.

\begin{verbatim}
cd /home/phenopi/phenotyping
sudo -u phenopi /home/phenopi/venvs/phenopi/bin/python \
-m scripts.scheduling.run_scheduler --help
\end{verbatim}

A successful installation should display the command-line help page. If the
command fails, check the repository path, virtual environment path, installed
dependencies, and scheduler module name.

\subsection{Installing the Scheduler Service}

Copy the provided service file to the systemd service directory and reload
systemd so that it detects the new service.

\begin{verbatim}
sudo cp /home/phenopi/phenotyping/deploy/phenopi-scheduler.service \
/etc/systemd/system/phenopi-scheduler.service
sudo systemctl daemon-reload
\end{verbatim}

Enable and start the service so that it starts automatically after boot.

\begin{verbatim}
sudo systemctl enable phenopi-scheduler.service
sudo systemctl start phenopi-scheduler.service
\end{verbatim}

\subsection{Checking the Service}

Check whether the scheduler service is running.

\begin{verbatim}
sudo systemctl status phenopi-scheduler.service
\end{verbatim}

The service logs can be viewed with \texttt{journalctl}.

\begin{verbatim}
sudo journalctl -u phenopi-scheduler.service -f
\end{verbatim}

If the service fails to start, first check that the paths in
\texttt{deploy/phenopi-scheduler.service} match the installation paths. In
particular, verify \texttt{WorkingDirectory}, \texttt{ExecStart}, and whether
the \texttt{phenopi} user has permission to read the repository and write to
the configured output directories.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 4
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Operation}

This chapter describes the standard operations for running automated image
collection. The goal of normal operation is to define an imaging schedule,
allow the system to automatically acquire images, and prepare the resulting
image dataset for downstream analysis.

During routine experiments, users do not directly interact with the camera.
Instead, experiments are controlled through a schedule file that defines when
images should be captured. The Phenopi scheduler runs continuously in the
background and executes the required captures automatically.

\subsection{Creating an experiment schedule}

Phenopi experiments are defined using a schedule configuration file. This file
describes the duration of the experiment and the times at which images should be
captured.

An example schedule configuration is shown below.

\begin{verbatim}
{
    "start_date": "2026-05-01",
    "num_days": 14,
    "capture_times": [
        "08:00",
        "08:30",
        "09:00",
        "09:30",
        "10:00"
    ]
}
\end{verbatim}

The \texttt{start\_date} field defines the first day of imaging. The
\texttt{num\_days} field determines the total experiment duration. The
\texttt{capture\_times} field specifies the time points at which images are
acquired each day.

Capture times should normally be restricted to the photoperiod of the plants.
For example, when plants are grown under a 08:00--20:00 light cycle, images
should only be scheduled during this period to avoid interrupting the dark
period.

\subsection{Generating the acquisition schedule}

Before an experiment starts, the schedule builder converts the experiment
configuration into a complete list of individual capture jobs.

Each scheduled capture receives an exact timestamp. This generated schedule
represents the complete experimental acquisition protocol and remains fixed
during the experiment.

This approach ensures that image acquisition is reproducible and that the exact
planned imaging times are known before the experiment begins.

\subsection{Running an experiment}

Once the acquisition service is active and a valid schedule has been created,
no further user interaction is required.

The scheduler continuously checks the current time against the experiment
schedule. When a capture time is reached, the scheduler starts a separate
camera process, acquires the image, saves the result, and records that the job
has been completed.

Each capture is independent. The camera is initialized, used for one image, and
then closed again. This reduces the risk of camera instability during
experiments lasting multiple days or weeks.

The scheduler keeps track of completed captures using a persistent state file.
If the Raspberry Pi restarts, previously completed images are not repeated and
the experiment continues from the remaining scheduled captures.

\subsection{Captured image output}

All acquired images are stored in the configured capture directory.

Images are saved using the filename format:

\begin{verbatim}
capture_YYYYMMDD_HHMMSS.jpg
\end{verbatim}

For example:

\begin{verbatim}
capture_20260504_153945.jpg
capture_20260504_154000.jpg
capture_20260504_154015.jpg
\end{verbatim}

The timestamp contained in the filename represents the acquisition time and is
used automatically by the downstream analysis pipeline. Users should therefore
avoid renaming captured images.

Multiple images can be acquired at each time point. These technical replicate
captures allow measurement repeatability to be quantified and help distinguish
true biological changes from image acquisition noise.

\subsection{Monitoring an active experiment}

During an experiment, the scheduler log can be inspected to verify normal
operation.

The service log can be viewed using:

\begin{verbatim}
journalctl -u phenopi-scheduler -f
\end{verbatim}

The log reports scheduled captures, successful acquisitions, and any errors
encountered during operation.

It is recommended to verify the first few captures of every experiment by
checking that images are correctly exposed, focused, and contain the complete
plant tray.

\subsection{After experiment completion}

After the final scheduled capture has completed, the image directory represents
the raw dataset for the experiment.

The capture directory should be preserved unchanged. All downstream Phenopi
analysis tools operate directly on these timestamped images and write processed
outputs to separate results directories.

Keeping the original images unchanged ensures that analyses can always be
repeated with updated parameters or newer versions of the analysis pipeline.

%-------------------------------------------------------------------------------
%    REFERENCES
%-------------------------------------------------------------------------------

% \newpage
% \printbibliography%

%-------------------------------------------------------------------------------
%    SUPPLEMENTAL
%-------------------------------------------------------------------------------

% \clearpage
% \input{startapp}
% \subsection{Supplemental figures and tables}

\end{document}
